Stellar Bars in Isolated Gas-Rich Spiral Galaxies Do Not Slow Down

### 2022-04-25 paper/ms.tex
Add Lars' edits

  Elongated bar-like features are ubiquitous, occurring at the centers of
  approximately two-thirds of spiral disk galaxies \cite{2000AJ....119..536E,
  2007ApJ...657..790M}. Due to gravitational interactions between the bar and
  the other components of galaxies, it is expected that angular momentum and
  matter will redistribute between galactic components over long (Gyr)
  timescales in galaxies hosting a bar \cite{1972MNRAS.157....1L,
  1984MNRAS.209..729T, 1985MNRAS.213..451W}. Previous work has overwhelmingly
  provided the expectation that, due to the interaction between the bar and
  the dark matter halo, the bar pattern will slow its rotation over time
  \cite{1992ApJ...400...80H, 2000ApJ...543..704D, 2002MNRAS.330...35A,
  2002ApJ...569L..83A, 2003MNRAS.341.1179A, 2003MNRAS.346..251O,
  2005MNRAS.363..991H, 2006ApJ...637..214M, 2007MNRAS.375..460W,
  2009ApJ...697..293D}. We have performed a simulation of a Milky Way-like
  galactic disk hosting a strong bar which includes a state-of-the-art model
  of the interstellar medium and a live dark matter halo. In this simulation
  the bar pattern does not slow down over time, and instead remains at a
  stable, constant rate of rotation. This behavior has been observed in
  previous simulations but its explanation has remained
  elusive.\cite{1993AA...268...65F, 2010ApJ...719.1470V}. We propose that the
  gas phase of the disk and the dark matter halo act in concert to stabilize
  the bar pattern and prevent it from slowing down or speeding up. This result
  naturally explains why nearly all observed bars rotate
  quickly\cite{2011MSAIS..18...23C, 2015AA...576A.102A, 2019MNRAS.482.1733G}
  and is especially relevant for our understanding of how the Milky Way
  arrived at its present day state.

@@

It has long been known that non-axisymmetric features in a stellar disk act to
redistribute mass and angular momentum.\cite{1972MNRAS.157....1L} The
interaction between a bar and a spheroid (either a dark matter halo or stellar
bulge) has received considerable interest.\cite{1984MNRAS.209..729T,
1985MNRAS.213..451W} A stellar disk in isolation seems to always generate a
bar instability,\cite{1971ApJ...168..343H} but the presence of a strong
spherical potential (e.g., from a stellar bulge or dark matter halo) acts to
stabilize the disk against the bar instability.\cite{1973ApJ...186..467O,
1976AJ.....81...30H} A number of numerical simulations have confirmed that a
live dark matter halo generically exerts a negative torque on the bar, causing
its pattern speed to decrease and its length and strength to
increase\cite{1992ApJ...400...80H, 2000ApJ...543..704D, 2002MNRAS.330...35A,
2002ApJ...569L..83A, 2003MNRAS.341.1179A, 2003MNRAS.346..251O,
2005MNRAS.363..991H, 2006ApJ...637..214M, 2007MNRAS.375..460W,
2009ApJ...697..293D}. Intuitively, the bar excites a wake of resonant material
in the dark matter halo in a manner similar to dynamical friction. This wake
lags and thus exerts a negative torque on the bar.

The preeminent expectation that the dark matter halo acts to slow the bar down
conflicts with observational estimates of bar pattern speeds. Bar rotation
rates are typically classified by the dimensionless ratio
\begin{equation}
\Rot = \RCR/R_b\textrm{,}
\end{equation}
where \RCR{} is the radius of corotation and $R_b$ is the length of the
bar.\footnote{The radius of corotation \RCR{} is defined for circular orbits
as the radius at which the orbital frequency is equal to the pattern speed,
$\Omega_p$, of a given non-axisymmetric feature. In a galaxy with a constant
circular velocity $V_c$, it is given by $\RCR = V_c / \Omega_p$.} Galaxies
with $\Rot < 1.4$ are considered ``fast rotators'' while galaxies with $\Rot >
1.4$ are considered ``slow rotators.''\cite{2000ApJ...543..704D} Galaxies with
$\Rot < 1$ are not thought to be stable.\cite{1980AA....81..198C}
Observational estimates of the pattern speeds of bars indicate that nearly all
galaxies have $1 < \Rot < 1.4$.\cite{2011MSAIS..18...23C, 2015AA...576A.102A,
2019MNRAS.482.1733G}

The role of gas on the evolution of the bar is less well-understood. Since the
gas phase typically only contributes about $10-20\%$ of the mass of a galaxy
at the present day, one might naively expect it to have a subdominant effect
on the bar. However, because gas is collisional, it can participate in
non-resonant angular momentum exchange with the bar.\cite{2011MNRAS.415.1027H}
Thus, numerical work has shown that the gas phase has an outsized influence on
the bar.\cite{2010ApJ...719.1470V, 2013MNRAS.429.1949A}

We have performed a simulation of a disk galaxy using the finite-volume
gravito-hydrodynamics code AREPO.\cite{2010MNRAS.401..791S} In AREPO, the
fluid is discretized as a moving Voronoi mesh. We use the additional physics
in the galaxy formation module Stars and MUltiphase Gas in GaLaxiEs (SMUGGLE)
\cite{2019MNRAS.489.4233M}. SMUGGLE is a comprehensive and self-consistent
galaxy formation model with a wide range of physical processes, including
radiative heating/cooling, star formation, and stellar feedback. More detailed
information on SMUGGLE is given in the Methods section. Our disk galaxy is a
modified version of the GALAKOS model\cite{2020ApJ...890..117D}, which
consists of a stellar disk, stellar bulge, and dark matter halo. After
$\sim2.5\,\textrm{Gyr}$ of evolution, the GALAKOS disk forms a bar consistent
with the Milky Way bar in terms of pattern speed and length
($\sim40\,\textrm{km}/\textrm{s}/\textrm{kpc}$ and $\sim4.5\,\textrm{kpc}$,
respectively). We modify this setup to include a gas phase, the details of
which is given in the Methods section.

A surface density projection of our simulations is shown in
Fig.~\ref{fig:overview}. The upper three panels show the disk in the N-body
run while the lower three panels show the disk in the SMUGGLE run. Each column
shows the disk $\sim1\,\textrm{Gyr}$ apart in time. There is a large qualitative
difference in the evolution of the bar pattern between the two runs. We see
that in the N-body case, the bar lengthens in time and grows in strength. In
the SMUGGLE case, the bar pattern retains a similar length and strength
between panels.

We show the time evolution of different bar properties in Fig.~\ref{fig:prop}.
In the upper panel, we show the pattern speed over time in the N-body (blue)
and SMUGGLE (orange) runs. The pattern speed in the N-body case slows down
while the pattern speed in the SMUGGLE case remains constant. The slowing down
of the pattern speed in the N-body case is consistent with a long line of
numerical research on bars in N-body simulations.\cite{1992ApJ...400...80H,
2000ApJ...543..704D, 2002MNRAS.330...35A, 2002ApJ...569L..83A,
2003MNRAS.341.1179A, 2003MNRAS.346..251O, 2005MNRAS.363..991H,
2006ApJ...637..214M, 2007MNRAS.375..460W, 2009ApJ...697..293D} However, in the
SMUGGLE case the pattern speed remains constant. After the first Gyr of
evolution, we find that the pattern speed increases by only $\sim10\%$ over
the next $4\,\textrm{Gyr}$, compared to a $\sim43\%$ decrease in the pattern
speed for the N-body run over the same interval. As we saw qualitatively in
Fig.~\ref{fig:overview}, the length of the bar in the N-body case grows over
time while it remains constant in the SMUGGLE case. This is also consistent
with previous numerical work, which found that bars tend to grow as they slow
down and the radius of corotation increases.\cite{2000ApJ...543..704D,
2003MNRAS.341.1179A}

@@

The fact that the dark matter halo in the SMUGGLE case exerts a smaller
positive torque on the bar can be understood in terms of the halo wake
mechanism. In the N-body case, halo material which is resonant with the bar
will form a wake which lags behind and exerts a negative torque on the bar,
which slows it down.\cite{1984MNRAS.209..729T, 1985MNRAS.213..451W,
1992ApJ...400...80H}\footnote{Since the bar is not a solid body, it is not
guaranteed that a negative torque will slow it down - e.g. a negative torque
could shred the bar, reducing its moment of inertia without changing its
pattern speed. However, the bar seems to empirically respond to a negative
torque induced by a halo wake by slowing down.} As the bar slows down, the
location of the resonances in phase space changes, allowing halo material
newly resonant with the bar to form a new wake. However, the gas is a reliable
source of positive torque on the bar, speeding the bar up. This stops the
resonance location from changing such that the halo cannot form a new wake,
arresting the process by which the halo can slow the bar down.

@@

The presence of the gas can arrest the process by which the dark matter halo
wake forms. However, this does not explain why the pattern speed in the
SMUGGLE case is nearly constant over several Gyr. Naively, it would be a
coincidence that the bar pattern speed remains constant in the SMUGGLE case,
resulting from a chance cancellation of the halo and gas torques. However, a
constant pattern speed in the presence of gas has been observed in a few
simulations of barred galaxies with gas.\cite{1993AA...268...65F,
2010ApJ...719.1470V} Previous work has argued this is due to the bar torquing
gas inwards, but no explanation has been given for why it might remain
constant.

@@

We performed two additional tests to explore how our findings depend on the
assumed model of the interstellar medium and the amount of gas present in the
system. First, we performed a simulation of the same disk but with a simpler
model of the interstellar medium\cite{2003MNRAS.339..289S}, closer to standard
methods used in cosmological simulations of galaxy formation. We find that the
pattern speed evolution is nearly the same in this case, and so conclude that
our result is not sensitive to the details of the model for the interstellar
medium. The details of this experiment is given in the Methods section.

Second, we performed a set of simulations in which we decreased the initial
gas mass of the disk. We essentially find that there is a certain critical gas
fraction (maybe gas surface density?) above which our proposed mechanism
operates. If the gas available is below this threshold, then the pattern speed
will decrease in a similar fashion to standard N-body simulations.
Importantly, arbitrarily increasing the gas fraction does not cause the
pattern speed to continuously increase. This is consistent with our proposed
equilibrium mechanism and with previous studies of barred galaxies which
varied the gas fraction.\cite{2010ApJ...719.1470V} The exact critical gas
fraction above which our mechanism operates is likely dependent on the
properties of the bar, disk, and dark matter halo -- further work is needed to
characterize how this gas fraction varies with such properties. Details on
this test in which we varied the gas mass of the disk is given in the Methods
section.

The implications of this finding are numerous. First, we naturally explain why
nearly all observed galaxies are fast rotators without requiring the inner
regions of dark matter halos to be underdense\cite{1998ApJ...493L...5D,
2000ApJ...543..704D} or introducing new physics.\cite{2021MNRAS.503.2833R,
2021MNRAS.508..926R} Second, we show that the role of gas is of paramount
importance in studies which attempt to uncover the nature of dark matter from
its effect of slowing down the bar.\cite{2021MNRAS.500.4710C,
2021MNRAS.505.2412C} Third, we provide an explanation for how the Milky Way's
bar could be both long-lived and a fast rotator, of which there is some
observational evidence.\cite{2019MNRAS.490.4740B} And finally, we complicate
the picture of radial mixing expected to sculpt the Milky Way's disk
\cite{2012MNRAS.420..913B, 2015ApJ...808..132H}, a process which relies upon
the pattern speed of the bar to change with time (though our work does not
alter expectations for radial mixing induced by spiral
arms\cite{2002MNRAS.336..785S}).

>>>>

  Elongated bar-like features are ubiquitous in galaxies, occurring at the centers of
  approximately two-thirds of spiral disks \cite{2000AJ....119..536E,
  2007ApJ...657..790M}.  Due to gravitational interactions between the bar and
  the other components of galaxies, it is expected that angular momentum and
  matter will redistribute over long (Gyr)
  timescales in barred galaxies \cite{1972MNRAS.157....1L,
  1984MNRAS.209..729T, 1985MNRAS.213..451W}. Previous work ignoring the gas phase of galaxies has conclusively 
  demonstrated that bars should slow their rotation over time 
  due to their interaction with dark matter halos
  \cite{1992ApJ...400...80H, 2000ApJ...543..704D, 2002MNRAS.330...35A,
  2002ApJ...569L..83A, 2003MNRAS.341.1179A, 2003MNRAS.346..251O,
  2005MNRAS.363..991H, 2006ApJ...637..214M, 2007MNRAS.375..460W,
  2009ApJ...697..293D}.  We have performed a simulation of a Milky Way-like
  galactic disk hosting a strong bar which includes a state-of-the-art model
  of the interstellar medium and a live dark matter halo. In this simulation
  the bar pattern does not slow down over time, and instead remains at a
  stable, constant rate of rotation. This behavior has been observed in
  previous simulations using more simplified models for the 
  interstellar gas but its explanation has remained
  elusive\cite{1993AA...268...65F, 2010ApJ...719.1470V}. We propose that the
  gas phase of the disk and the dark matter halo act in concert to stabilize
  the bar pattern speed and prevent a bar from slowing down or speeding up. This result
  naturally explains why nearly all observed bars rotate
  rapidly\cite{2011MSAIS..18...23C, 2015AA...576A.102A, 2019MNRAS.482.1733G}
  and is especially relevant for our understanding of how the Milky Way
  arrived at its present state.

@@

It has long been known that non-axisymmetric features in a stellar disk act to
redistribute mass and angular momentum.\cite{1972MNRAS.157....1L} The
interaction between a bar and a spheroid (either a dark matter halo or stellar
bulge) has received considerable interest.\cite{1984MNRAS.209..729T,
1985MNRAS.213..451W} Stellar disks in isolation are prone to developing bars, \cite{1971ApJ...168..343H} but the presence of a strong
spherical potential (e.g., from a stellar bulge or dark matter halo) acts to
stabilize the disk against bar formation.\cite{1973ApJ...186..467O,
1976AJ.....81...30H} Many numerical simulations have confirmed that a
live dark matter halo generically exerts a negative torque on a bar, causing
its pattern speed to decrease and its length and strength to
increase\cite{1992ApJ...400...80H, 2000ApJ...543..704D, 2002MNRAS.330...35A,
2002ApJ...569L..83A, 2003MNRAS.341.1179A, 2003MNRAS.346..251O,
2005MNRAS.363..991H, 2006ApJ...637..214M, 2007MNRAS.375..460W,
2009ApJ...697..293D}. Intuitively, the bar excites a wake of resonant material
in a dark matter halo in a manner akin to dynamical friction. This wake
lags a bar and thus exerts a negative torque on it.

The expectation that a dark matter halo acts to slow down a bar is in
conflict with observational estimates of bar pattern speeds. Bar rotation
rates are typically classified by the dimensionless ratio
\begin{equation}
\Rot = \RCR/R_b\textrm{,}
\end{equation}
where \RCR{} is the radius of corotation and $R_b$ is the length of the
bar.\footnote{The radius of corotation \RCR{} is defined for circular orbits
as the radius at which the orbital frequency is equal to the pattern speed,
$\Omega_p$, of a given non-axisymmetric feature. In a galaxy with a constant
circular velocity $V_c$, it is given by $\RCR = V_c / \Omega_p$.} Galaxies
with $\Rot < 1.4$ are considered ``fast rotators'' while galaxies with $\Rot >
1.4$ are considered ``slow rotators.''\cite{2000ApJ...543..704D} Galaxies with
$\Rot < 1$ are not thought to be stable.\cite{1980AA....81..198C}
Observational estimates of the pattern speeds of bars indicate that nearly all
galaxies have $1 < \Rot < 1.4$.\cite{2011MSAIS..18...23C, 2015AA...576A.102A,
2019MNRAS.482.1733G}

The role of gas on the evolution of the bar is less well-understood. Since the
gas phase typically only contributes about $10-20\%$ of the mass of a galaxy
at the present day, one might naively expect it to have a subdominant effect
on the bar. However, because gas is collisional, it can participate in
non-resonant angular momentum exchange with the bar.\cite{2011MNRAS.415.1027H}
Thus, numerical work has shown that the gas phase can have a stronger influence on
a bar than its contribution to the mass of a galaxy would suggest.\cite{2010ApJ...719.1470V, 2013MNRAS.429.1949A}

We have performed a simulation of a disk galaxy using the finite-volume
gravito-hydrodynamics code AREPO.\cite{2010MNRAS.401..791S} In AREPO, the
fluid is discretized as a moving Voronoi mesh. We use the additional physics
in the galaxy formation module Stars and MUltiphase Gas in GaLaxiEs (SMUGGLE)
\cite{2019MNRAS.489.4233M}. SMUGGLE is a comprehensive and self-consistent
galaxy formation model with a wide range of physical processes, including
radiative heating/cooling, star formation, and stellar feedback. More detailed
information on SMUGGLE is given in the Methods section. Our disk galaxy is a
modified version of the GALAKOS model\cite{2020ApJ...890..117D}, which
consists of a stellar disk, stellar bulge, and dark matter halo. After
$\sim2.5\,\textrm{Gyr}$ of evolution, the GALAKOS disk forms a bar consistent
with the Milky Way bar in terms of pattern speed and length
($\sim40\,\textrm{km}/\textrm{s}/\textrm{kpc}$ and $\sim4.5\,\textrm{kpc}$,
respectively). We modify this setup to include a gas phase, the details of
which are provided in the Methods section.

A surface density projection of our simulations is shown in
Fig.~\ref{fig:overview}. The upper three panels show the disk in the N-body
run while the lower three panels show the disk in the SMUGGLE run. Each column
shows the disk $\sim1\,\textrm{Gyr}$ apart in time. There is a large qualitative
difference in the evolution of the bar pattern between the two runs. We see
that in the N-body case, the bar lengthens in time and grows in strength. In
the SMUGGLE case, the bar retains a similar length and strength
between panels.

We show the time evolution of different bar properties in Fig.~\ref{fig:prop}.
In the upper panel, we show the pattern speed over time in the N-body (blue)
and SMUGGLE (orange) runs. The pattern speed in the N-body case slows down
while the pattern speed in the SMUGGLE case remains roughly constant. The slowing down
of the pattern speed in the N-body case is consistent with a long line of
numerical research on bars in N-body simulations.\cite{1992ApJ...400...80H,
2000ApJ...543..704D, 2002MNRAS.330...35A, 2002ApJ...569L..83A,
2003MNRAS.341.1179A, 2003MNRAS.346..251O, 2005MNRAS.363..991H,
2006ApJ...637..214M, 2007MNRAS.375..460W, 2009ApJ...697..293D} However, in the
SMUGGLE case the pattern speed remains constant. After the first Gyr of
evolution, we find that the pattern speed increases by only $\sim10\%$ over
the next $4\,\textrm{Gyr}$, compared to a $\sim43\%$ decrease in the pattern
speed for the N-body run over the same interval. As we saw qualitatively in
Fig.~\ref{fig:overview}, the length of the bar in the N-body case grows over
time while it remains roughly constant in the SMUGGLE case. This is also consistent
with previous numerical work, which found that bars tend to grow as they slow
down and the radius of corotation increases.\cite{2000ApJ...543..704D,
2003MNRAS.341.1179A}

@@

The fact that the dark matter halo in the SMUGGLE case exerts a smaller
positive torque on the bar can be understood in terms of the halo wake
mechanism. In the N-body case, halo material which is resonant with the bar
will form a wake which lags behind and exerts a negative torque on the bar,
which slows it down.\cite{1984MNRAS.209..729T, 1985MNRAS.213..451W,
1992ApJ...400...80H}\footnote{Since the bar is not a solid body, it is not
guaranteed that a negative torque will slow it down - e.g. a negative torque
could shred the bar, reducing its moment of inertia without changing its
pattern speed. However, the bar seems to empirically respond to a negative
torque induced by a halo wake by slowing down.} As the bar slows down, the
location of the resonances in phase space changes, allowing halo material
newly resonant with the bar to participate in forming the  wake. However, 
the gas is a reliable
source of positive torque on the bar, speeding the bar up. This stops the
resonance location from changing such that the halo cannot reinforce a wake,
arresting the process by which the halo can slow the bar down.

@@

The presence of the gas can arrest the process by which additional material in
the dark matter halo can contribute to a
wake. However, this does not explain why the pattern speed in the
SMUGGLE case is nearly constant over several Gyr. Naively, it would be a
coincidence that the bar pattern speed remains constant in the SMUGGLE case,
resulting from a chance cancellation of the halo and gas torques. However, a
constant pattern speed in the presence of gas has been observed in a few
simulations of barred galaxies with gas.\cite{1993AA...268...65F,
2010ApJ...719.1470V} Previous work has argued this is due to the bar torquing
gas inwards, but no explanation has been given for why it might remain
constant.

@@

We performed two additional tests to explore how our findings depend on the
assumed model of the interstellar medium and the amount of gas present in the
system. First, we performed a simulation of the same disk but with a simpler
model of the interstellar medium\cite{2003MNRAS.339..289S}, closer to standard
methods used in cosmological simulations of galaxy formation. We find that the
pattern speed evolution is nearly the same in this case, and so conclude that
our result is not sensitive to the details of the model for the interstellar
medium. The details of this experiment are given in the Methods section.

Second, we performed a set of simulations in which we decreased the initial
gas mass of the disk. We essentially find that there is a certain critical gas
fraction (maybe gas surface density?) above which our proposed mechanism
operates. If the gas available is below this threshold, then the pattern speed
will decrease in a similar fashion to standard N-body simulations.
Importantly, arbitrarily increasing the gas fraction does not cause the
pattern speed to continuously increase. This is consistent with our proposed
equilibrium mechanism and with previous studies of barred galaxies which
varied the gas fraction.\cite{2010ApJ...719.1470V} The exact critical gas
fraction above which our mechanism operates is likely dependent on the
properties of the bar, disk, and dark matter halo -- further work is needed to
characterize how this gas fraction varies with such properties. Details on
this test in which we varied the gas mass of the disk are given in the Methods
section.

The implications of our findings are numerous. First, we naturally explain why
nearly all observed galaxies are fast rotators without requiring the inner
regions of dark matter halos to be underdense\cite{1998ApJ...493L...5D,
2000ApJ...543..704D} or requiring new physics.\cite{2021MNRAS.503.2833R,
2021MNRAS.508..926R} Second, we show that the role of gas is of paramount
importance in studies which attempt to uncover the nature of dark matter from
its effect of slowing down the bar.\cite{2021MNRAS.500.4710C,
2021MNRAS.505.2412C} Third, we provide an explanation for how the Milky Way's
bar could be both long-lived and a fast rotator, of which there is some
observational evidence.\cite{2019MNRAS.490.4740B} And finally, we complicate
the picture of radial mixing expected to sculpt the Milky Way's disk
\cite{2012MNRAS.420..913B, 2015ApJ...808..132H}, a process which relies upon
the pattern speed of the bar to change with time (though our work does not
alter expectations for radial mixing induced by spiral
arms\cite{2002MNRAS.336..785S}).
